Rodent trapping studies can improve our understanding of the hazards of zoonotic pathogen spillover and novel zoonotic pathogen emergence.

### 2022-05-24 reports/manuscript_v4.Rmd
update for v4

Rodents are important reservoirs of known zoonotic infectious diseases and novel zoonotic pathogens. Rodent reservoirs are globally distributed with West Africa containing multiple host species of zoonotic infectious diseases. Ongoing anthropogenic land use change alters the composition of rodent species assemblages which can alter the hazard of zoonotic disease spillover and emergence. Current global datasets to understand these hazards suffer from biases. Here, we synthesise data from 127 rodent trapping studies including over 72,000 individual rodent detections to understand the range and presence of important rodent reservoir species to quantify the bias in global datasets. We identify the rodent trapping studies are biased towards human dominated landscapes with important sampling biases across West Africa. We find that these rodent trapping studies can complement current global datasets on rodent population distributions which can minimise some of these observed biases. For four regionally important zoonotic infectious diseases we identify host-pathogen associations not included in a comprehensive global dataset. Rodent trapping studies can further enrich these resources through geolocating these associations and producing measures of pathogen prevalence within hosts. Finally, we find that for most of these diseases the proportion of a rodent hosts range that has been sampled remains small. Further rodent trapping studies are required to understand the true hazard of zoonotic spillover of these diseases across West Africa. Our results highlight that understanding rodent sampling bias is important when assessing the risk for zoonotic spillover events from rodents. The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data. We anticipate that this dataset can support the production of more complete understanding of the hazard of zoonotic spillover events.

@@

Zoonotic infectious disease spillover events into human populations (animal-to-human transmission) can have important health and economic consequences at both the global and regional level (e.g. SARS-CoV-2 and Ebola virus disease) [@judson_zoonoses_2021]. The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropenic pressure driven by increased human populations, urbanisation, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. West Africa is a nexus for these processes and a hotspot for both the risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@grace_mapping_2012; @han_global_2016; @allen_global_2017]. Two taxa - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_global_2016]. Specifically, West Africa has been identified as region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. Currently within West Africa rodents are involved in the transmission of multiple zoonotic diseases with large burdens on human health, including, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021]. Rodent species with "fast" life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].

Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are based on global datasets of host species distributions and host-pathogen associations (e.g., IUCN, GBIF, GIDEON) which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations), resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching global datasets and mitigating against their inherent biases [@bovendorp_optimising_2017]. Rodent trapping provides a useful method to describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in global datasets. Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of known zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021]. However, data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.

Here, we synthesise rodent trapping studies conducted across West African from a search of literature between 1964-2019. First, we investigate the geographic biases in the rodent trapping dataset in relation to human population density and land use classification. Second, we compare this to global host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions. Third, we compare identified host-pathogen associations with global datasets (CLOVER) to understand discrepencies in rodent host-pathogen associations and report the proportion positive for pathogens of interest. Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of undersampling of pathogens within their host ranges. We expect that rodent trapping studies will provide an important additional source of high-resolution data that can be used to enrich global datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa. We find that rodent trapping studies identify greater geographic ranges of important rodent hosts of zoonotic diseases than are reported from global datasets while also identifying locations and habitats in which further sampling is required. We also identify host-pathogen associations that are not reported in global host-pathogen association data, while identifying systematic undersampling for these pathogens across their host ranges.

@@

We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date limits were set: (1) Rodent OR Rodent trap* AND (2) West Africa (or the individual countries).

We searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms. Searches were run on 2021-03-01, returning studies conducted between 1964-2019.

We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation. We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English. One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria. At each stage, a random subset (10%) was reviewed by a second reviewer.

We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.). Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo]. Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest. We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.

We extracted GPS locations for the most precise location presented, converting to decimal degrees as required. For studies not using standardised habitat recording (e.g., IUCN Habitat Classification Scheme (Version 3.1)), the explicit description of the habitat in which the trap was placed was extracted. For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.

We mapped genus and species names to those in the Global Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]. We extracted information on the detection, non-detection and number of trapped individuals.

We extracted data on all pathogens assayed. The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used. If studies reported indeterminate results this was noted. Pathogens were identified to species level, however, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used.

### Spatial covariates

Boundaries of countries and level-2 administrative regions were obtained from GADM 4.0.4 for the study region [@database_of_global_administrative_areas_gadm_2022]. Land cover classification in 2005, the median year in which trapping activity occurred, was obtained from Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022]. Proportion of land cover type within a level-2 administrative region was calculated. Human population density for 2005 was obtained from SEDAC at ~ 1km resolution, with median population density within each region extracted [@socioeconomic_data_and_applications_center_gridded_2021].

### Rodent species ranges and presence

Rodent species distributions were obtained as shapefiles from the IUCN redlist, shapefiles were cropped to the study region [@iucn_iucn_2021]. Rodent presence locations were obtained from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021].

### Host pathogen associations

CLOVER a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw 2020) was released in 2021 and represents a comprehensive global repository of host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi [@gibb_data_2021]. Data were obtained from an archived version of the dataset [@gibb_viralemergenceclover_2021]. The dataset was limited to species trapped within our included studies, no spatial subsetting was possible.

## Included studies

We identified 4,282 relevant citations, with 126 rodent trapping studies included (Supplementary Table 2.). The earliest trapping studies were conducted in 1964, with increasing numbers of studies being performed annually. The median year of first trapping activity was 2005, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.). Studies were conducted in 14 West African countries with no studies reported from Gambia or Togo, at 1,331 sites (Figure 1A.). A minority (31, 25%) of studies trapped at a single study site, with 46 (37%) trapping at between two and five sites, the remaining 49 studies (38%) trapped at between six and 93 study sites. There were 581,426 reported trap nights, with 384,983 trap nights imputed for studies with no recording of trapping effort (Figure 1B.).

![Figure 1: A) The location of trapping sites in West Africa. No sites were recorded from Togo or Gambia. There is important heterogeneity observed in the coverage of each country by trap night (colour) and location of sites. For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso. B) The distribution of trap nights performed at each study site on a log scale with a median of 244 trap nights (IQR 115-494) performed at each site](`r here("figures", "Figure_1_updated.png")`)

Within the included studies trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared to other habitats (3%, IQR 1-9%). In total 73,164 small mammals were trapped with 62,574 (85%) identified to species level and 7,760 (11%) identified to genus. The majority of the 140 identified species were Rodentia (110, 76%) and Soricomorpha (30, 21%), of which Muridae (82, 73%) were the most common family.

62 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens. Most studies tested for a single microorganism (39, 62%). The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (27, 43%), *Borrelia sp.* (8, 13%), *Bartonella sp.* (4, 6%) and *Toxoplasma gondii* (4, 6%). Most studies used Polymerase Chain Reaction to detect microorganisms (32, 52%), with fewer studies using molecular based tests (11, 18%) or histological or direct visualisation assays (8, 13%). From 32,014 individual rodent samples we produced 339 host-pathogen pairs. With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.

Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data has been made available online [@simons_rodents_2022].

@@

We used the GPS coordinates of trapping locations and the number of trap nights to calculate trapping effort (trap night density) within level-2 administrative regions in West Africa. We imputed trap nights for studies with missing records using the number of trapped rodents and median trap success rate from studies reporting trapping effort stratified by habitat type. We calculated the proportion of land cover classes within each of these level-2 administrative, comparing surveyed regions to non-surveyed regions.

To better understand the potential bias associated with studies for zoonotic disease spillover or novel zoonotic pathogen emergence we limited further analyses to zoonotic pathogen studies only. We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association between number of trap nights, proportion of urban, cropland and forest land classification and human population density [@pedersen_hierarchical_2019]. The model was constructed in *mgcv* using the R statistical programme [@wood_generalized_2017; @r_core_team_r_2020]. Selection of the most parsimonious model was based on the adjusted-R^2^ and Deviance explained for each model. We performed sensitivity analysis by removing sites with imputed trapping effort. The final selected model structure was specified as:

$$
\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie((Urban) + (Forest) + (Longitude * Latitude))}
\end{equation}
$$

The model identified locations of high and low trapping effort relative to land use classification, with proportion urban land very highly correlated with human population density (*r* = 0.835, 95% C.I. 0.82-0.85, *p* < 0.001). Model inference was limited to regions around observed trap sites.

### What is the difference between global datasets rodent host distributions and distributions from trapping studies?

We assessed the taxonomic coverage of the trapped taxa and for the seven species with the most trap locations (*M. natalensis*, *R. rattus*, *M. erythroleucus*, *M. musculus*, *A. niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped the detection and non-detection of rodent species and compared this to detection data obtained from GBIF and species ranges from IUCN [@gbif_the_global_biodiversity_information_facility_occurrence_2021; @iucn_iucn_2021]. We identified rodent locations from GBIF and the rodent trapping studies within and outside the IUCN expected range. We located these to level-2 administrative regions to calculate the discrepencies between observed detections and non-detections with IUCN distributions.

### Are rodent trapping derived host-pathogen associations present in global zoonoses datasets?

For studies investigating rodent zoonoses we produced host-pathogen pairs. For each host-pathogen pair we compared our records with those available in CLOVER, reporting concordance and discordance. For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER. We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.

@@

For the four pathogens sampled at the most locations (Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) we identify their potential host species. For the five species with most spatially complete sampling we associate their sampling locations with level-2 administrative regions and compare this to their IUCN range and rodent trapping detection range to calculate a measure sampling coverage.

@@

Trap sites were situated in 273 of 1,414 (19.3%) level-2 administrative regions in 14 West African nations. The areas with highest trap night density included the capital and large cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), Ghana (Accra),  Côte d'Ivoire (Abidjan), and Benin (Cotonou). Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 1A.).

Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats. However, urban habitats were comparitively undersampled (Supplementary Figure 2.).

>>>>

NEEDS UPDATING

@@

There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@Bernstein]. The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. The impact of endemic zoonoses meanwhile remains underestimated [@Maudlin]. Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@Halliday; @Molyneux; @judson_zoonoses_2021]. West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016]. Two taxonomic groups - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015]. Specifically, West Africa has been identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times (<4 days) further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021]. Rodent species with “fast” life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].

Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@Bowler; @Boakes]. These sampling biases can gravely distort produced species distribution models [@Beck]. Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@Gibb_clover]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations, endemic zoonoses), potentially resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching consolidated datasets and providing additional data to minimise inherent biases [@bovendorp_optimising_2017]. Rodent trapping studies describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in consolidated datasets. Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of endemic zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify emerging and novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021]. However, individual level data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.

Here, we synthesise rodent trapping studies conducted across West Africa from a search of literature between 1964-2022. First, we use this dataset to investigate the geographic sampling biases in relation to human population density and land use classification. Second, we compare this to curated host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions. Third, we compare identified host-pathogen associations with a consolidated dataset (CLOVER) to explore discrepancies in rodent host-pathogen associations and report the proportion of positive assays for pathogens of interest. Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of sparse sampling of pathogens within their host ranges. We expect that rodent trapping studies provide an important additional source of high-resolution data that can be used to enrich available consolidated datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.

@@

To identify relevant literature, we conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as exploded keywords: (1) Rodent OR Rodent trap* AND (2) West Africa, no date limits were set. We also searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms. Searches were run on 2022-05-01, and returned studies conducted between 1964-2021.

For further analysis, we included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, and v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation. We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English. One author screened titles, abstracts and full texts against the inclusion and exclusion criteria. At each stage; title screening, abstract screening and full text review, a random subset (10%) was reviewed by a second author.

We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.). Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo on finalised]. We identified the aims of included studies, for example, whether it was conducted as a survey of small mammal species or specifically to assess the risk of zoonotic disease spillover. To understand potential limitations in the representativeness of rodent trapping we extracted data on study methodology, such as, the number of trap nights, the type of traps used and whether the study attempted to estimate abundance. For studies not reporting number of trap nights we used imputation based on the number of trapped individuals, stratified by the habitat type from which they were obtained. This was performed by multiplying the total number of trapped individuals within that study site by the median trap success for study sites with the same reported habitat type. Stratification was used as trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared with other habitats (3%, IQR 1-9%)

The method of taxonomic identification applied within a study was recorded, reported species identification was assumed to be accurate. The number of individuals of these species or genera was extracted with species and genus names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021] for internal consistency. We expanded species detection and non-detection records by explicitly specifying non-detection at a trap site if a species was recorded as detected at other trapping locations within the same study.

Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented. GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@GeoNames] based on place-names and maps presented in the study. All locations were converted to decimal degrees for internal consistency. The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity. The habitats of trapping sites were mapped to the IUCN Habitat Classification Scheme (Version 3.1). For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.

For relevant studies we extracted data on all microorganisms and zoonotic pathogens tested and the method used (e.g., molecular or serological diagnosis). Where assays were able to identify the microorganism to species level this was recorded, non-specific assays higher order attribution was used (e.g. to family level). We recorded the species of rodent host tested, the number of individuals tested and the number of positive and negative results. For studies reporting summary results all testing data were extracted, this may introduce double counting of individual rodents, for example, if a single rodent was tested using both molecular and serological assays. Where studies reported indeterminate results, these were also recorded.

We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted. We identified 4,692 relevant citations, with 127 rodent trapping studies included (Supplementary Table 2.). The earliest trapping studies were conducted in 1964, with a trend of increasing numbers of studies being performed annually since 2000. The median year of first trapping activity was 2007, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.). Studies were conducted in 14 West African countries, with no studies reported from The Gambia or Togo, at 1,611 trap sites (Figure 1A.).

Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success. A total of 942,629 trap nights were estimated (Figure 1B.). A minority of studies trapped at a single study site (30, 24%), with 46 (36%) trapping at between two and five sites, the remaining 51 studies (40%) trapped at between six and 93 study sites.

In total 76,275 small mammals were trapped with 65,628 (90%) identified to species level and 7,439 (10%)  identified to genus, with the remaining classified to higher taxonomic level. The majority of the 132 identified species were Rodentia (102, 78%), of which Muridae (73, 72%) were the most common family. Soricomorpha were the second most identified order of small mammals (28, 21%). 57 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens. Most studies tested for a single microorganism (48, 84%). The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (21, 37%), *Borrelia sp.* (9, 16%), *Bartonella sp.* (4, 7%) and *Toxoplasma gondii* (4, 7%). Most studies used Polymerase Chain Reaction (PCR) to detect microorganisms (37, 65%), with fewer studies using serology-based tests (11, 19%) or histological or direct visualisation assays (11, 21%). From 32,920 individual rodent samples we produced 351 host-pathogen pairs. With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms. Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data is available as supplementary material [@simons_rodents_2022].

![Figure 1: A) The location of trapping sites in West Africa. No sites were recorded from Togo or The Gambia. Heterogeneity is observed in the coverage of each country by trap night (colour) and location of sites. For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso. B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.png")`)

@@

To investigate the risk to human health from endemic zoonoses and novel pathogen emergence using synthesised data from rodent trapping data understanding of sampling biases is vital. Sampling of rodents has not been performed randomly across space in West Africa (Figure 1.) and at varying intensity, we aimed quantify this bias as the relationship between trapping effort (measured as trap night density (TN density)), human population density and land use type. We define this as relative trapping effort bias.

To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@Database of Global Administrative Areas]. We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above. The sf package in the R statistical language was used to handle shapefiles and geographic data [@Pebesma]. Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area. For studies not reporting trap nights imputation was used as previously described. Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@SEDAC].  Median population density was then calculated for each level-2 administrative region. Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@COPERNICUS]. The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.

We investigated the role of these covariates on the spatial bias of TN density by comparing to a spatial model (equation 1), as follows:

$$
\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie((Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{2}
\text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{3}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{4}
\text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}
\end{equation}

\begin{equation}
\tag{5}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Proportion urban) + (Longitude * Latitude))}
\end{equation}
$$

We formulated these models as a Generalised Additive Models (GAM) incorporating a spatial interaction term (longitude by latitude) [@pedersen_hierarchical_2019]. These models investigated the association between TN density, proportion of urban, cropland, tree cover and human population density. The models were constructed in the *mgcv* package using the R programming language [@wood_generalized_2017; @r_core_team_r_2020]. Selection of the most parsimonious model was based on Deviance explained and the Akaike information criterion for each model. We performed sensitivity analysis by removing sites with imputed trapping effort and performing a pixel-based analysis.

### What is the difference in rodent host distributions between curated datasets and rodent trapping studies?

Curated datasets (i.e., IUCN and GBIF) are commonly used in assessments of biodiversity in response to land use and climate change in addition to being used to explore zoonotic pathogen host distributions. To understand any biases introduced from these sources into these analyses we aimed to assess the concordance of curated rodent host distributions from IUCN and GBIF with observed rodent presence and absence from rodent trapping studies. We obtained rodent species distribution maps as shapefiles from the IUCN red list and translated these to a ~20km resolution raster [@iucn_iucn_2021]. Distributions were cropped to the study region for globally distributed rodent species. We obtained rodent presence locations from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021]. Presence locations were associated to cells of raster with a ~20km resolution produced for the study region.

We assessed the taxonomic coverage of the trapped taxa and for the seven species with the most trap locations (*M. natalensis*, *R. rattus*, *M. erythroleucus*, *M. musculus*, *A. niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped detection and non-detection, associating each location with a ~20km resolution raster.  For these seven species we calculated the area of the IUCN expected range. We calculated the percentage of this range covered by presence detections in GBIF, this process was repeated for presence detections from rodent trapping studies. For both GBIF data and trapping studies we calculated the area of detections outside of the IUCN expected range. For rodent trapping studies only, we were able to calculate the area of non-detections within the IUCN expected area. Finally, we calculated the combined area of detection when combining GBIF and rodent trapping study data.

### Are rodent trapping derived host-pathogen associations present in a consolidated zoonoses dataset?

Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations. To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@Gibb_data_proliferation]. CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi. An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@Gibb_clover_data].

We produced host-pathogen pairs from rodent trapping studies investigating zoonoses. For each host and pathogen, with both identified to species level, we compared our host-pathogen pair network with one produced using CLOVER, reporting concordance and discordance. We describe these networks and report differences in network composition. For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER. We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.

@@

Sampling of zoonotic pathogens may also be biased within their hosts ranges, which may hinder inference across the entire range of the species. To quantify this, we focused on the five pathogen groups and species sampled at the most locations (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*). We geolocated individuals of host species tested for these pathogens and calculated the area sampled. We adopt a similar approach to above by mapping sampled hosts to ~20km pixels and calculating the proportion of their IUCN range in which sampling has occurred. We also present the area covered of all areas in which the host has been detected in this study to produce a measure of relative completeness of sampling within the included rodent trapping studies.

Data and code implemented in the R statistical language has been made available alongside this manuscript to reproduce these analyses [ref_zenodo].

@@

We found a non-random spatial distribution of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa. Trap sites were situated in 256 of 1,450 (17.6%) level-2 administrative regions in 14 West African nations. The regions with highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou). Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Figure 1A.).

Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats. However, urban habitats were comparatively undersampled (Supplementary Figure 2.).
